---
title: "Midterm Project"
author: "Ravi Brenner, Cameron Chesbrough, Wayne Monical"
date: "2025-03-25"
output: pdf_document
---





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Library Load
```{r}
library(caret)
library(tidymodels)
library(vtable)
library(corrplot)
library(patchwork)
library(splines)
library(mgcv)
library(pdp)
library(earth)
library(tidyverse)
library(ggplot2)
library(glmnet)
```

Loading data
```{r}
load("dat1.RData")
load("dat2.RData")
```

## Exploratory Data Analysis

Go from labels to variable names
```{r}
dat1 <- dat1 |>
  mutate(gender = factor(gender, levels = c(0,1),
                         labels = c("Female","Male")),
         race = factor(race, levels = c(1,2,3,4),
                       labels = c("White","Asian","Black","Hispanic")),
         smoking = factor(smoking, levels = c(0,1,2),
                          labels = c("Never smoked","Former smoker","Current smoker")),
         diabetes = factor(diabetes, levels = c(0,1),
                           labels = c("No","Yes")),
         hypertension = factor(hypertension, levels = c(0,1),
                               labels = c("No","Yes")),
         ) |>
  dplyr::select(-id)

dat2 <- dat2 |>
  mutate(gender = factor(gender, levels = c(0,1),
                         labels = c("Female","Male")),
         race = factor(race, levels = c(1,2,3,4),
                       labels = c("White","Asian","Black","Hispanic")),
         smoking = factor(smoking, levels = c(0,1,2),
                          labels = c("Never smoked","Former smoker","Current smoker")),
         diabetes = factor(diabetes, levels = c(0,1),
                           labels = c("No","Yes")),
         hypertension = factor(hypertension, levels = c(0,1),
                               labels = c("No","Yes")),
         ) |>
  dplyr::select(-id)
```

Use featureplot from caret to plot the training data
```{r}
theme1 <- trellis.par.get()
theme1$plot.symbol$col = rgb(.2, .2, .2, .4)
theme1$plot.symbol$pch = 16
theme1$plot.line$col = rgb(1, 0, 0, .7)
theme1$plot.line$lwd <- 2
trellis.par.set(theme1)

featurePlot(x = dat1 |> dplyr::select(-log_antibody,
                                      -where(is.factor)),
            y = dat1$log_antibody,
            type = c("p","smooth"))

dat1 |> 
  dplyr::select(log_antibody,gender,race, smoking,diabetes, hypertension) |>
  pivot_longer(cols = 2:6,
               names_to = "variable",
               values_to = "value") |>
  ggplot(aes(y = value, x = log_antibody)) + 
  geom_boxplot() +
  facet_wrap(.~variable)
```

```{r}

antibody_hist = hist(dat1$log_antibody)
antibody_scatter = plot(x = dat1$time, y = dat1$log_antibody)
summ_table = sumtable(dat1, out = 'return')
continuous = dat1[c(1,5:7,10:12)]
correlations = cor(continuous)
corr_plot = corrplot(correlations)

report_table = sumtable(dat1, out = 'kable')
report_table


```

Looking at the second dataset
```{r}
antibody_hist = hist(dat1$log_antibody)
antibody_hist_data2 = hist(dat2$log_antibody)

summ_table_data2 = sumtable(dat2, out = 'return')
```


## Model Training

Creating the design Matrix
```{r}
load('dat1.RData')

design_matrix = 
  dat1 |>
  mutate(
    race_asian = as.numeric(race == 2),
    race_black = as.numeric(race == 3),
    race_hispanic = as.numeric(race == 4),
    smoking_former = as.numeric(smoking == 1),
    smoking_current = as.numeric(smoking ==2)
  ) %>% 
  select(
    age, gender, race_asian, race_black, race_hispanic, 
    smoking_former, smoking_current, height, weight,
    bmi, diabetes, hypertension, SBP, LDL, time, 
    log_antibody
  )
```
  
Specify X and Y for model training
```{r}
y = design_matrix$log_antibody
x = select(design_matrix, -log_antibody) %>% 
  as.matrix()
```

Specify CV Procedure
```{r}
ctrl <- trainControl(method = "cv", number = 10)
```

Linear Regression
```{r}
set.seed(1)
model.linear =
  train(x = x,
        y = y,
        method = "lm", 
        metric = "RMSE",
        trControl = ctrl)

hist(model.linear$resample$RMSE)
```

```{r}
coef(model.linear$finalModel)
```


Elastic Net
```{r}
set.seed(1)
model.elastic_net =
  train(x = x,
        y = y,
        method = "glmnet", 
        metric = "RMSE",
        trControl = ctrl,
        tuneGrid = expand.grid(.alpha = seq(0,1, length = 21),
                               .lambda = exp(seq(6,0, length = 100))))

hist(model.elastic_net$resample$RMSE)
```


```{r}
plot(model.elastic_net)
```



GAM
```{r}
set.seed(1)
model.gam =
  train(x = x,
        y = y,
        method = "gam", 
        metric = "RMSE",
        trControl = ctrl)
```

```{r}
plot(model.gam)
```


MARS
```{r}
mars_grid = 
  expand.grid(degree = 1:3,
              nprune = 2:15)

set.seed(1)
model.mars = 
  train(x, y,
        method = "earth",
        tuneGrid = mars_grid,
        trControl = ctrl)

ggplot(model.mars) + 
  labs('MARS Model Evaluation')
```


## Comparing Cross Validated RMSE

```{r}
model.RMSE= 
  rbind(
  data.frame(
    model = 'Linear',
    RMSE = model.linear$resample$RMSE
  ),
  data.frame(
    model = 'Elastic Net',
    RMSE = model.elastic_net$resample$RMSE),
  data.frame(
    model = 'GAM',
    RMSE = model.gam$resample$RMSE
  ),
  data.frame(
    model = 'MARS',
    RMSE = model.mars$resample$RMSE
  )
)
```


```{r}
model.RMSE %>% 
  group_by(model) %>% 
  summarize(mean(RMSE))
```


```{r}

model.RMSE |>
  ggplot(aes(x = model, y = RMSE)) +
  geom_violin()+
  stat_summary(
    fun = "mean",
               geom = "point",
               color = "red")+
  labs(title = "Model RMSE Comparison", xlab = "Model")
```


## Model Evaluation

Clean new data set
```{r}
load('dat2.RData')

design_matrix2 = 
  dat2 |>
  mutate(
    race_asian = as.numeric(race == 2),
    race_black = as.numeric(race == 3),
    race_hispanic = as.numeric(race == 4),
    smoking_former = as.numeric(smoking == 1),
    smoking_current = as.numeric(smoking ==2)
  ) %>% 
  select(
    age, gender, race_asian, race_black, race_hispanic, 
    smoking_former, smoking_current, height, weight,
    bmi, diabetes, hypertension, SBP, LDL, time, 
    log_antibody
  )

x2 = as.matrix(dplyr::select(design_matrix2, -log_antibody))
y2 = design_matrix2$log_antibody
```


```{r}
head(x)
```
```{r}
head(x2)
```


Make predictions and get test set RMSE
```{r}
pred = predict(model.gam, x2)

dat2_rmse = sqrt(mean((pred - y2)^2))

dat2_rmse
```
